---
title: 'R&R: Peer-graded Assignment: Course Project 1'
author: "kamal pradhan"
date: "15 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data
The data for this assignment was downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Setting environment and loading libraries
setting **echo = TRUE** that someone else will be able to read the code and loading libraries **ggplot2** which will be used for plotting.


```{r e_setup}
#importing and configuring knitr
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')

#loading other required libraries
library(ggplot2) 
library(lattice)
```

##Loading and preprocessing the data
Assuming that the file is already extracted and present in the working directory we directly load the file.
```{r load_preprocess}
#loading the csv file
activity_data<-read.csv('activity.csv',header = TRUE, sep = ",",stringsAsFactors = FALSE)

#formatting the date column
activity_data$date<-as.Date(activity_data$date, format="%Y-%m-%d")

#converting the interval column to factor type
activity_data$interval<-as.factor(activity_data$interval)
```

##What is mean total number of steps taken per day?

####Calculating the total number of steps taken per day
here the calculation for steps per day is done using tapply. here we also ignore the the ***missing values***.

```{r step_day}
#calculating steps per date
steps_day <- tapply(activity_data$steps, activity_data$date, FUN = sum, na.rm = TRUE)
```

####Histogram of the total number of steps taken each day
```{r hist_plot}
#histogram plot
hist(steps_day,col = "green",xlab = "Total Steps", ylab = "Frequency",main = "Steps Taken Each Day")

```

####Mean and Median of the total number of steps taken per day
```{r mean_median}
#mean and median of steps taken per day
mean_steps   <- mean(steps_day)
median_steps <- median(steps_day)
```

The mean and median of total no of steps taken per day are  **`r format(mean_steps,digits = 8)`** and **`r format(median_steps,digits = 8)`** respectively.

##What is the average daily activity pattern?


####Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r t_plot}
#calculating average number of steps taken in 5-minute interval 
mean_activity <- aggregate(activity_data$steps, by = list(interval=activity_data$interval), FUN = mean, na.rm = TRUE)

#renaming columns
names(mean_activity) <- c("interval", "average")

#Time series plot
plot(as.integer(mean_activity$interval),
     mean_activity$average,
     type="l", 
     col="green", 
     xlab="Interval in mins", 
     ylab="Average steps", 
     main="Time-series plot of Average steps in 5 min Interval")
```

####Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_steps}
#calculating interval containing max steps
max_interval<-mean_activity[which.max(mean_activity$average),1]
```
The 5-minute interval that contains maximum steps is **`r format(max_interval,digits = 8)`**.

##Imputing missing values

####Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s)
```{r t_missing}
#calculating missing values
missing_values <- is.na(activity_data$steps)
missing_values<-table(missing_values)[2]
```

Total missing value is **`r format(missing_values,digits = 8)`**.

####Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r d_impute}
#finding NA positions
pos_na <- which(is.na(activity_data$steps))

# imputing using mean
imp_mean <- rep(mean(activity_data$steps, na.rm=TRUE), times=length(pos_na))
```

####Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r fill_missing}
#filling missing values
activity_data[pos_na, "steps"] <- imp_mean
```

####Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r new_hist}
#calculating steps per date
steps_day <- tapply(activity_data$steps, activity_data$date, FUN = sum, na.rm = TRUE)

#histogram plot
hist(steps_day,col = "green",xlab = "Total Steps", ylab = "Frequency",main = "Steps Taken Each Day")

#mean and median of steps taken per day
mean_steps   <- mean(steps_day)
median_steps <- median(steps_day)
```
The new mean and median of total no of steps taken per day are  **`r format(mean_steps,digits = 8)`** and **`r format(median_steps,digits = 8)`** respectively.

Mean and Median values are increased after the imputation. Mean and Median also tends to be same after the imputation. 

##Are there differences in activity patterns between weekdays and weekends?

####Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r week_day}
#creating a weekday vector
week_days <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')

#specifying week day or weekend based on the weekday vector
activity_data$daytype <- c('weekend', 'weekday')[(weekdays(activity_data$date) %in% week_days)+1L]

#sample dataframe after weekday calculation
head(activity_data)
```

####Make a panel plot containing a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r week_dayplot}
#calculating average steps taken based on weekends and weekdays
activity_data_mean <- aggregate(activity_data$steps,by=list(activity_data$daytype, activity_data$interval), mean)

#renamicng the columns
names(activity_data_mean) <- c("daytype", "interval", "Average")

#plotting the graph for weekend and week days
xyplot(Average ~ interval | daytype, activity_data_mean, 
       type="l", 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
````
